When two-point correlation function fails to describe a structure...

### 2024-03-06 paper.tex
Capitalize probabilities

\textcolor{red}{How to cite a theorem in the paper?}
Another useful result is that for large enough $d$ the probability
of $f$ having factorization $\Phi \tilde{f}$ where $\tilde{f}$ is an irreducible
polynomial and $\Phi$ is a product of cyclotomic polynomials is at least
\begin{equation}
  p = 1 - 2\exp(-c(\log d)^{\beta - 2})
  \label{eq:prob-cyc-1}
\end{equation}
for some $c > 0$ and $\beta > 3$ if $\zeta_K$ has no zeros $\rho$ with
$|1 - \rho| < (\log d)^\alpha/d$, $\alpha > \beta$ for all $K = \mathbb{Q}(a)$
for any root $a$ of $f$.

Now we will prove a simple lemma which connects these results with collisions in
correlation functions.
\begin{theorem}
  \label{the:sequence}
  Let $A$ be a sufficiently long random sequence of 0's and 1's where each
  element is chosen randomly and independently with equal probability $1/2$ with
  exception of the first and the last elements which are 1's. Then, with
  probability at least
  \begin{equation*}
    p = 1 - 2\exp(-c(\log d)^\beta)
  \end{equation*}
  for some $c > 0$, $\beta > 1$ the sequence $A$ of the length $d+1$ does not
  have non-trivial collisions in the two-point correlation function.
\end{theorem}
\begin{proof}
  Let $f(x)$ be an image of $A$ under Z-transform and a change of variables
  $x = z^{-1}$. With probability at least $p$ it factors in $\mathbb{Z}[x]$ as
  $\Phi \tilde{f}$, where $\Phi$ is a product of cyclotomic polynomials and
  $\tilde{f}$ is irreducible. All cyclotomic polynomials with exception of
  $\Phi_1 = x - 1$ are palindromes. $\Phi_1$ does not divide $f$ because
  $f(1) = \sum\limits_{k=0}^{d-1} f_k \ge 2$. Therefore, $f$ has at most one
  non-palindrome factor and hence $A$ does not have non-trivial collisions.
\end{proof}

@@

Now we get our main result:
\begin{theorem}
  Let $A$ be a 2-dimensional $D_1 \times D_2$ array of 0's and 1's where each
  element is chosen randomly and independently with probability $1/2$ with
  exception of two elements $A[0, 0]$ and $A[D_1 - 1, 0]$, which are
  1's. Suppose the Riemann hypothesis for the Dedekind zeta function $\zeta_K$
  holds for all fields $K = \mathbb{Q}(a)$ where $a$ is any root of $f_x(x)$ and
  $f_x(x)$ is an image of the slice $A[0:D_1 -1, 0]$ under Z-transform and a
  change in variables $x = z^{-1}$. Then $A$ does not have non-trivial
  collisions in the two-point correlation function with probability at least
  \begin{equation}
    p = 1 - \sqrt{\frac{2}{\pi (D_1 - 1)}}
    \label{eq:main-result}
  \end{equation}
\end{theorem}
\begin{proof}
  Let $f(x, y)$ be an image of $A$ under Z-transform and a change of variables
  $x = z_1^{-1}, y = z_2^{-1}$ with $\deg_x f = d$, where $d = D_1 - 1$. Let
  $\phi f = f \mod y$ be a ``partial evaluation'' homomorphism from
  $\mathbb{Z}[x, y]$ to $\mathbb{Z}[x]$. Oviously,
  $f_x(x) = \phi f = f(x,0)$. Under conditions of the theorem $f(0) = 1$ and
  $\deg f_x = d$.  According to \cref{eq:prob-irr} we have
  \begin{equation*}
    p\{\text{$f_x$ is irreducible}\} = 1 - \sqrt{\frac{2}{\pi d}} + O(d^{-1})
  \end{equation*}
  Now we need to prove a similar result for $f$.

  We have three possible cases of factorization of $f$:
  \begin{itemize}
  \item $f$ is irreducible.
  \item $f$ factors as $f(x, y) = g(x)h(y)$. In this case
    $\phi h = h(0) = const$ and \cref{lem:factors} cannot be applied. According
    to distribution of coefficients in $f$ the probability of this case is
    $p_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$. For any fixed integer $D_2 > 1$
    we have $p_{split} = o(D_1^{-1})$. For $D_2 = 1$ we have $f = f_x$ and
    \cref{eq:prob-irr} is applicable directly to $f$.
  \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions on
    the coefficients $\deg_x g > 0$ and $\deg_x h > 0$, therefore
    $\phi g \ne const$ and $\phi h \ne const$. We have
    $f_x = (\phi g) (\phi h)$. Applying \cref{lem:factors}, we can say that
    $f$ is irreducible with probability at least \cref{eq:main-result}.
  \end{itemize}
  Hence $f$ is irreducible with probability at least \cref{eq:main-result}. If
  $f$ is irreducible, there is no two or more non-palindrome factors of $f$,
  hence $A$ does not have non-trivial collisions.
\end{proof}
Surely, $D_1$ can be replaced with $D_2$ in \cref{eq:main-result} if
$A[0, D_2 - 1] = 1$. This theorem also can be proved for multidimensional arrays
by recursively applying the same argument.

>>>>

\textcolor{red}{How to cite a theorem in the paper?}
Another useful result is that for large enough $d$ the probability
of $f$ having factorization $\Phi \tilde{f}$ where $\tilde{f}$ is an irreducible
polynomial and $\Phi$ is a product of cyclotomic polynomials is at least
\begin{equation}
  P = 1 - 2\exp(-c(\log d)^{\beta - 2})
  \label{eq:prob-cyc-1}
\end{equation}
for some $c > 0$ and $\beta > 3$ if $\zeta_K$ has no zeros $\rho$ with
$|1 - \rho| < (\log d)^\alpha/d$, $\alpha > \beta$ for all $K = \mathbb{Q}(a)$
for any root $a$ of $f$.

Now we will prove a simple lemma which connects these results with collisions in
correlation functions.
\begin{theorem}
  \label{the:sequence}
  Let $A$ be a sufficiently long random sequence of 0's and 1's where each
  element is chosen randomly and independently with equal probability $1/2$ with
  exception of the first and the last elements which are 1's. Then, with
  probability at least
  \begin{equation*}
    P = 1 - 2\exp(-c(\log d)^\beta)
  \end{equation*}
  for some $c > 0$, $\beta > 1$ the sequence $A$ of the length $d+1$ does not
  have non-trivial collisions in the two-point correlation function.
\end{theorem}
\begin{proof}
  Let $f(x)$ be an image of $A$ under Z-transform and a change of variables
  $x = z^{-1}$. With probability at least $P$ it factors in $\mathbb{Z}[x]$ as
  $\Phi \tilde{f}$, where $\Phi$ is a product of cyclotomic polynomials and
  $\tilde{f}$ is irreducible. All cyclotomic polynomials with exception of
  $\Phi_1 = x - 1$ are palindromes. $\Phi_1$ does not divide $f$ because
  $f(1) = \sum\limits_{k=0}^{d-1} f_k \ge 2$. Therefore, $f$ has at most one
  non-palindrome factor and hence $A$ does not have non-trivial collisions.
\end{proof}

@@

Now we get our main result:
\begin{theorem}
  Let $A$ be a 2-dimensional $D_1 \times D_2$ array of 0's and 1's where each
  element is chosen randomly and independently with probability $1/2$ with
  exception of two elements $A[0, 0]$ and $A[D_1 - 1, 0]$, which are
  1's. Suppose the Riemann hypothesis for the Dedekind zeta function $\zeta_K$
  holds for all fields $K = \mathbb{Q}(a)$ where $a$ is any root of $f_x(x)$ and
  $f_x(x)$ is an image of the slice $A[0:D_1 -1, 0]$ under Z-transform and a
  change in variables $x = z^{-1}$. Then $A$ does not have non-trivial
  collisions in the two-point correlation function with probability at least
  \begin{equation}
    P = 1 - \sqrt{\frac{2}{\pi (D_1 - 1)}}
    \label{eq:main-result}
  \end{equation}
\end{theorem}
\begin{proof}
  Let $f(x, y)$ be an image of $A$ under Z-transform and a change of variables
  $x = z_1^{-1}, y = z_2^{-1}$ with $\deg_x f = d$, where $d = D_1 - 1$. Let
  $\phi f = f \mod y$ be a ``partial evaluation'' homomorphism from
  $\mathbb{Z}[x, y]$ to $\mathbb{Z}[x]$. Oviously,
  $f_x(x) = \phi f = f(x,0)$. Under conditions of the theorem $f(0) = 1$ and
  $\deg f_x = d$.  According to \cref{eq:prob-irr} we have
  \begin{equation*}
    P\{\text{$f_x$ is irreducible}\} = 1 - \sqrt{\frac{2}{\pi d}} + O(d^{-1})
  \end{equation*}
  Now we need to prove a similar result for $f$.

  We have three possible cases of factorization of $f$:
  \begin{itemize}
  \item $f$ is irreducible.
  \item $f$ factors as $f(x, y) = g(x)h(y)$. In this case
    $\phi h = h(0) = const$ and \cref{lem:factors} cannot be applied. According
    to distribution of coefficients in $f$ the probability of this case is
    $P_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$. For any fixed integer $D_2 > 1$
    we have $P_{split} = o(D_1^{-1})$. For $D_2 = 1$ we have $f = f_x$ and
    \cref{eq:prob-irr} is applicable directly to $f$.
  \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions on
    the coefficients $\deg_x g > 0$ and $\deg_x h > 0$, therefore
    $\phi g \ne const$ and $\phi h \ne const$. We have
    $f_x = (\phi g) (\phi h)$. Applying \cref{lem:factors}, we can say that
    $f$ is irreducible with probability at least \cref{eq:main-result}.
  \end{itemize}
  Hence $f$ is irreducible with probability at least \cref{eq:main-result}. If
  $f$ is irreducible, there is no two or more non-palindrome factors of $f$,
  hence $A$ does not have non-trivial collisions.
\end{proof}
Surely, $D_1$ can be replaced with $D_2$ in \cref{eq:main-result} if
$A[0, D_2 - 1] = 1$. This theorem also can be proved for multidimensional arrays
by recursively applying the same argument.
